Simulating High-Dimensional Multivariate Data

### 2020-10-30 040-monte-carlo-evaluation.Rmd
run bivariate sims rev2

*Bivariate Normal*.
Let's simulate a bivariate normal and check our correlation matching performance as N increases.
Here we have BVN( $\mu_1 = \mu_2 = 10, \rho_{type}$ ).
We vary $\rho$ across the entire possible range of correlations for each correlation type.

```{r ch040-biNormal, echo = FALSE, eval = TRUE}
mom_norm <- function(x) {
  m <- mean(x)
  s <- sd(x)
  list(mean = m, sd = s)
}

mu <- 10
sigma <- 1
margins <- alist(
    qnorm(mean = mu, sd = sigma),
    qnorm(mean = mu, sd = sigma)
)

type <- c("pearson", "spearman", "kendall")
## cores <- as.integer( parallel::detectCores() - 1 )
## n <- c(1e3, 1e4, 1e5)
n <- c(1e1, 1e2, 1e3)
adjustForDiscrete <- c(FALSE)

eps <- 1e-2
## grid_steps <- 100
grid_steps <- 10
sim_pars <- expand.grid(type = type, cores = cores, n = n,
                        stringsAsFactors = FALSE,
                        adjustForDiscrete = adjustForDiscrete)

res <- data.frame()
dir.create("./results/sims_norm", showWarnings = TRUE, recursive = TRUE)
for (i in 1:nrow(sim_pars)) {
    type <- sim_pars$type[i]
    cores <- sim_pars$cores[i]
    n <- sim_pars$n[i]
    adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
    tmp_bounds <- cor_bounds(margins, type = type)
    cor_lo <- tmp_bounds$lower[1,2] + eps
    cor_hi <- tmp_bounds$upper[1,2] - eps
    cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
    ## rho = cor_seq[1]
    for (rho in cor_seq) {
        Rho <- matrix(rho, 2, 2)
        diag(Rho) <- 1.0
        ## Rho <- convertCor( rho = Rho, from = type, to = 'pearson' )
        time_data <- system.time({
            ## x <- rmvn(n = n, mu = rep( mu, 2 ), sigma = Rho)
            ## use rvec to test our software
            x <- rvec(n = n,
                      rho = Rho,
                      margins = margins,
                      type = type,
                      cores = cores)
        })
        ## Save the sims in case
        id <- paste0(
            "d", 2,
            "-N", n,
            "-c", cores,
            "-r", rho,
            "-Cor", type,
            "-adj", as.character(adjustForDiscrete),
            "-dev", "1CORE",
            "-lib", "bigsimr"
        )
        ## save sims in their own subdirectory
        saveRDS(x, file = paste0("./results/sims_norm/", id, ".rds"))

        ## Estimate statistics
        Rho_hat <- cor_fast(x, method = type)
        rho_hat <- Rho_hat[1, 2]
        norm_args_hat <- mom_norm(x[,1])
        mu_hat <- norm_args_hat$mean
        sigma_hat <- norm_args_hat$sd

        ## Save the results
        res <- rbind(res, data.frame(
                              method = "bigsimr",
                              device = "CPU",
                              type = type,
                              cores = cores,
                              margins = "norm",
                              adjustForDiscrete = adjustForDiscrete,
                              d = 2,
                              N = n,
                              rho = rho,
                              rho_hat = rho_hat,
                              mean = mu,
                              sd = sigma,
                              mean_hat = mu_hat,
                              sd_hat = sigma_hat,
                              sim_time = unname(time_data["elapsed"])
                          ))
    }
}
res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
saveRDS(object = res, "./results/norm_sims.rds")
```

Figure \@ref(fig:ch040-biNormPlot) `bigsimr` recovers the Pearson specified correlations for MVN.

```{r ch040-biNormPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for MVN."}
dat <- readRDS("./results/norm_sims.rds")
dat %>%
    filter(cores == cores) %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    ## ggplot(aes(rho, rho_hat)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ + N) + theme_bw()
    facet_wrap(~ type + N) + theme_bw()

## ggsave('fig/plot-biNormPlot.pdf')
```

*Bivariate Gamma*. 
Similarly, let's check the performance for a non-symmetric continuous distribution: a standard (rate =1) bivariate gamma.
Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
We vary $\rho$ across the entire possible range of correlations for each correlation type.

```{r ch040-biGamma, echo = FALSE, eval = TRUE}
mom_gamma <- function(x) {
  m <- mean(x)
  s <- sd(x)
  list(shape = m^2 / s^2, rate = m / s^2)
}

shape <- 10
rate <- 1
margins <- alist(
    qgamma(shape = shape, rate = rate),
    qgamma(shape = shape, rate = rate)
)

type <- c("pearson", "spearman", "kendall")
## cores <- c(1L)
## n <- c(1e3, 1e4, 1e5)
n <- c(1e1, 1e2, 1e3)
adjustForDiscrete <- c(FALSE)

eps <- 1e-2
## grid_steps <- 100
grid_steps <- 10
sim_pars <- expand.grid(type = type, cores = cores, n = n,
                        stringsAsFactors = FALSE,
                        adjustForDiscrete = adjustForDiscrete)

dir.create("./results/sims_gamma", showWarnings = TRUE, recursive = TRUE)
res <- data.frame()
for (i in 1:nrow(sim_pars)) {

@@

        ## Save the sims in case
        id <- paste0(
            "d", 2,
            "-N", n,
            "-c", cores,
            "-r", rho,
            "-Cor", type,
            "-adj", as.character(adjustForDiscrete),
            "-dev", "CPU",
            "-lib", "bigsimr"
        )
        saveRDS(x, file = paste0("./results/sims_gamma/", id, ".rds"))

@@

        ## Save the results
        res <- rbind(res, data.frame(
                              method = "bigsimr",
                              device = "CPU",
                              type = type,
                              cores = cores,
                              margins = "gamma",
                              adjustForDiscrete = adjustForDiscrete,
                              d = 2,
                              N = n,
                              rho = rho,
                              rho_hat = rho_hat,
                              shape = shape,
                              rate = rate,
                              shape_hat = shape_hat,
                              rate_hat = rate_hat,
                              sim_time = unname(time_data["elapsed"])
                          ))
    }
}
res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
saveRDS(object = res, "./results/sims_gamma.rds")
```

Figure \@ref(fig:ch040-biGammaPlot) `bigsimr` recovers the Pearson specified correlations for bivariate gamma.

```{r ch040-biGammaPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
dat <- readRDS("./results/sims_gamma.rds")
dat %>%
    filter(cores == cores) %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    ## ggplot(aes(rho, rho_hat)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ + N) + theme_bw()
    facet_wrap(~ type + N) + theme_bw()

## ggsave('fig/plot-biGammaPlot.pdf')
```

*Bivariate Negative Binomial*. 
Let's check the performance for a discrete distribution: a bivariate negative binomial.
Here we have Bivariate Negative Binomial ( $prob_1 = prob_2 = 0.5, size_1 = size_2 = 4,\rho_{type}$ ).
We vary $\rho$ across the entire possible range of correlations for each correlation type.

```{r ch040-biNegBin, echo = FALSE, eval = TRUE}
mom_nbinom <- function(x) {
  m <- mean(x)
  s <- sd(x)
  list(size = m^2 / (s^2 - m), prob = m / s^2)
}

size <- 10
prob <- 0.1
margins <- alist(
    qnbinom(size = size, prob = prob),
    qnbinom(size = size, prob = prob)
)

type <- c("pearson", "spearman", "kendall")
## cores <- cores
## n <- c(1e3, 1e4, 1e5)
n <- c(1e1, 1e2, 1e3)
adjustForDiscrete <- c(FALSE)

eps <- 1e-2
## grid_steps <- 100
grid_steps <- 10
sim_pars <- expand.grid(type = type, cores = cores, n = n,
                        stringsAsFactors = FALSE,
                        adjustForDiscrete = adjustForDiscrete)

dir.create("./results/nbinom_sims", showWarnings = TRUE, recursive = TRUE)
res <- data.frame()
for (i in 1:nrow(sim_pars)) {
    type <- sim_pars$type[i]
    cores <- sim_pars$cores[i]
    n <- sim_pars$n[i]
    adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
    ## AGS. 17 Oct 2020. cor_bounds failed.
    ## I'll create a minimal working example and open an issue
    ## tmp_bounds <- cor_bounds(margins, type = type)
    ## cor_lo <- tmp_bounds$lower[1,2] + eps
    ## cor_hi <- tmp_bounds$upper[1,2] - eps
    cor_lo <- -0.75
    cor_hi <- 0.95
    cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)

@@

        ## Save the sims in case
        id <- paste0(
            "d", 2,
            "-N", n,
            "-c", cores,
            "-r", rho,
            "-Cor", type,
            "-adj", as.character(adjustForDiscrete),
            "-dev", "CPU",
            "-lib", "bigsimr"
        )
        saveRDS(x, file = paste0("./results/nbinom_sims/", id, ".rds"))

@@

        ## Save the results
        res <- rbind(res, data.frame(
                              method = "bigsimr",
                              device = "CPU",
                              type = type,
                              cores = cores,
                              margins = "nbinom",
                              adjustForDiscrete = adjustForDiscrete,
                              d = 2,
                              N = n,
                              rho = rho,
                              rho_hat = rho_hat,
                              size = size,
                              prob = prob,
                              size_hat = size_hat,
                              prob_hat = prob_hat,
                              sim_time = unname(time_data["elapsed"])
                          ))
    }
}
res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
saveRDS(object = res, "./results/nbinom_sims.rds")
```

Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial.

```{r ch040-biNegBinPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
dat <- readRDS("./results/nbinom_sims.rds")
dat %>%
    filter(cores == cores) %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    ## ggplot(aes(rho, rho_hat)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ + N) + theme_bw()
    facet_wrap(~ type + N) + theme_bw()
## ggsave('fig/plot-biNegBinPlot.pdf')
```

>>>>

```{r ch040-biNormal, echo = FALSE, eval = TRUE}
mu <- 0
sigma <- 1
if ( !file.exists( 'results/biNorm_sims.rds' ) )  {
    mom_norm <- function(x) {
        m <- mean(x)
        s <- sd(x)
        list(mean = m, sd = s)
    }
    mu <- 0
    sigma <- 1
    margins <- alist(
        qnorm(mean = mu, sd = sigma),
        qnorm(mean = mu, sd = sigma)
    )
    type <- c("pearson", "spearman", "kendall")
    cores <- as.integer( parallel::detectCores() - 1 )
    n <- c(1e3, 1e4, 1e5)
    ## n <- c(1e1, 1e2, 1e3)
    adjustForDiscrete <- c(FALSE)
    eps <- 1e-2
    grid_steps <- 100
    ## grid_steps <- 10
    sim_pars <- expand.grid(type = type, cores = cores, n = n,
                            stringsAsFactors = FALSE,
                            adjustForDiscrete = adjustForDiscrete)
    res <- data.frame()
    dir.create("./results/biNorm_sims", showWarnings = TRUE, recursive = TRUE)
    for (i in 1:nrow(sim_pars)) {
        type <- sim_pars$type[i]
        cores <- sim_pars$cores[i]
        n <- sim_pars$n[i]
        adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
        tmp_bounds <- cor_bounds(margins, type = type)
        cor_lo <- tmp_bounds$lower[1,2] + eps
        cor_hi <- tmp_bounds$upper[1,2] - eps
        cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)
        ## rho = cor_seq[1]
        for (rho in cor_seq) {
            Rho <- matrix(rho, 2, 2)
            diag(Rho) <- 1.0
            ## Rho <- convertCor( rho = Rho, from = type, to = 'pearson' )
            time_data <- system.time({
                ## x <- rmvn(n = n, mu = rep( mu, 2 ), sigma = Rho)
                ## use rvec to test our software
                x <- rvec(n = n,
                          rho = Rho,
                          margins = margins,
                          type = type,
                          cores = cores)
            })
            ## Save the sims in case
            id <- paste0(
                "d", 2,
                "-N", n,
                "-c", cores,
                "-r", rho,
                "-Cor", type,
                "-adj", as.character(adjustForDiscrete),
                "-dev", "1CORE",
                "-lib", "bigsimr"
            )
            ## save sims in their own subdirectory
            saveRDS(x, file = paste0("./results/biNorm_sims/", id, ".rds"))
            ## Estimate statistics
            Rho_hat <- cor_fast(x, method = type)
            rho_hat <- Rho_hat[1, 2]
            norm_args_hat <- mom_norm(x[,1])
            mu_hat <- norm_args_hat$mean
            sigma_hat <- norm_args_hat$sd
            ## Save the results
            res <- rbind(res, data.frame(
                                  method = "bigsimr",
                                  device = "CPU",
                                  type = type,
                                  cores = cores,
                                  margins = "norm",
                                  adjustForDiscrete = adjustForDiscrete,
                                  d = 2,
                                  N = n,
                                  rho = rho,
                                  rho_hat = rho_hat,
                                  mean = mu,
                                  sd = sigma,
                                  mean_hat = mu_hat,
                                  sd_hat = sigma_hat,
                                  sim_time = unname(time_data["elapsed"])
                              ))
        }
    }
    res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
    saveRDS(object = res, "./results/biNorm_sims.rds")
}
```

*Bivariate Normal*.
Let's simulate a bivariate normal and check our correlation matching performance as N increases.
Standard bivariate normal 
Here we have BVN( $\mu_1 = \mu_2 = 0, \rho_{type}$ ).
We vary $\rho$ across the entire possible range of correlations for each correlation type.
Figure \@ref(fig:ch040-biNormPlot) `bigsimr` recovers the Pearson specified correlations for MVN.

```{r ch040-biNormPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for MVN."}
dat <- readRDS("./results/biNorm_sims.rds")
dat %>%
    filter(cores == cores) %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    ## ggplot(aes(rho, rho_hat)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ + N) + theme_bw()
    facet_wrap(~ type + N) + theme_bw()
## ggsave('fig/plot-biNormPlot.pdf')
```

```{r ch040-biGamma, echo = FALSE, eval = TRUE}
shape <- 10
rate <- 1
if ( !file.exists( 'results/biGamma_sims.rds' ) )  {
    mom_gamma <- function(x) {
        m <- mean(x)
        s <- sd(x)
        list(shape = m^2 / s^2, rate = m / s^2)
    }
    margins <- alist(
        qgamma(shape = shape, rate = rate),
        qgamma(shape = shape, rate = rate)
    )

    type <- c("pearson", "spearman", "kendall")
    ## cores <- c(1L)
    cores <- as.integer( parallel::detectCores() - 1 )
    n <- c(1e3, 1e4, 1e5)
    ## n <- c(1e1, 1e2, 1e3)
    adjustForDiscrete <- c(FALSE)

    eps <- 1e-2
    grid_steps <- 100
    ## grid_steps <- 10
    sim_pars <- expand.grid(type = type, cores = cores, n = n,
                            stringsAsFactors = FALSE,
                            adjustForDiscrete = adjustForDiscrete)

    dir.create("./results/biGamma_sims", showWarnings = TRUE, recursive = TRUE)
    res <- data.frame()
    for (i in 1:nrow(sim_pars)) {

@@

            ## Save the sims in case
            id <- paste0(
                "d", 2,
                "-N", n,
                "-c", cores,
                "-r", rho,
                "-Cor", type,
                "-adj", as.character(adjustForDiscrete),
                "-dev", "CPU",
                "-lib", "bigsimr"
            )
            saveRDS(x, file = paste0("./results/biGamma_sims/", id, ".rds"))

@@

            ## Save the results
            res <- rbind(res, data.frame(
                                  method = "bigsimr",
                                  device = "CPU",
                                  type = type,
                                  cores = cores,
                                  margins = "gamma",
                                  adjustForDiscrete = adjustForDiscrete,
                                  d = 2,
                                  N = n,
                                  rho = rho,
                                  rho_hat = rho_hat,
                                  shape = shape,
                                  rate = rate,
                                  shape_hat = shape_hat,
                                  rate_hat = rate_hat,
                                  sim_time = unname(time_data["elapsed"])
                              ))
        }
    }
    res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
    saveRDS(object = res, "./results/biGamma_sims.rds")
}
```

*Bivariate Gamma*. 
Similarly, let's check the performance for a non-symmetric continuous distribution: a standard (rate =1) bivariate gamma.
Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
We vary $\rho$ across the entire possible range of correlations for each correlation type.
Figure \@ref(fig:ch040-biGammaPlot) `bigsimr` recovers the Pearson specified correlations for bivariate gamma.

```{r ch040-biGammaPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width= 8, fig.align='center', fig.cap = "`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
dat <- readRDS("./results/biGamma_sims.rds")
dat %>%
    filter(cores == cores) %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    ## ggplot(aes(rho, rho_hat)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ + N) + theme_bw()
    facet_wrap(~ type + N) + theme_bw()
## ggsave('fig/plot-biGammaPlot.pdf')
```

```{r ch040-biNegBin, echo = FALSE, eval = TRUE}
size <- 4
prob <- 3e-04
if ( !file.exists( 'results/biNB_sims.rds' ) )  {
    mom_nbinom <- function(x) {
        m <- mean(x)
        s <- sd(x)
        list(size = m^2 / (s^2 - m), prob = m / s^2)
    }

    margins <- alist(
        qnbinom(size = size, prob = prob),
        qnbinom(size = size, prob = prob)
    )

    type <- c("pearson", "spearman", "kendall")
    cores <- as.integer( parallel::detectCores() - 1 )
    ## cores <- cores
    n <- c(1e3, 1e4, 1e5)
    ## n <- c(1e1, 1e2, 1e3)
    adjustForDiscrete <- c(FALSE)

    eps <- 1e-2
    grid_steps <- 100
    ## grid_steps <- 10
    sim_pars <- expand.grid(type = type, cores = cores, n = n,
                            stringsAsFactors = FALSE,
                            adjustForDiscrete = adjustForDiscrete)

    dir.create("./results/biNB_sims", showWarnings = TRUE, recursive = TRUE)
    res <- data.frame()
    for (i in 1:nrow(sim_pars)) {
        type <- sim_pars$type[i]
        cores <- sim_pars$cores[i]
        n <- sim_pars$n[i]
        adjustForDiscrete <- sim_pars$adjustForDiscrete[i]
        ## AGS. 17 Oct 2020. cor_bounds failed.
        ## I'll create a minimal working example and open an issue
        ## tmp_bounds <- cor_bounds(margins, type = type)
        ## cor_lo <- tmp_bounds$lower[1,2] + eps
        ## cor_hi <- tmp_bounds$upper[1,2] - eps
        cor_lo <- -0.75
        cor_hi <- 0.95
        cor_seq <- seq(cor_lo, cor_hi, length.out = grid_steps)

@@

            ## Save the sims in case
            id <- paste0(
                "d", 2,
                "-N", n,
                "-c", cores,
                "-r", rho,
                "-Cor", type,
                "-adj", as.character(adjustForDiscrete),
                "-dev", "CPU",
                "-lib", "bigsimr"
            )
            saveRDS(x, file = paste0("./results/biNB_sims/", id, ".rds"))

@@

            ## Save the results
            res <- rbind(res, data.frame(
                                  method = "bigsimr",
                                  device = "CPU",
                                  type = type,
                                  cores = cores,
                                  margins = "nbinom",
                                  adjustForDiscrete = adjustForDiscrete,
                                  d = 2,
                                  N = n,
                                  rho = rho,
                                  rho_hat = rho_hat,
                                  size = size,
                                  prob = prob,
                                  size_hat = size_hat,
                                  prob_hat = prob_hat,
                                  sim_time = unname(time_data["elapsed"])
                              ))
        }
    }
    res$type <- factor(res$type, levels = c("pearson",  "spearman", "kendall" ) )
    saveRDS(object = res, "./results/biNB_sims.rds")
}
```

*Bivariate Negative Binomial*. 
Let's check the performance for a discrete distribution: a bivariate negative binomial.
Here we have Bivariate Negative Binomial ( $prob_1 = prob_2 = 0.5, size_1 = size_2 = 4,\rho_{type}$ ).
We vary $\rho$ across the entire possible range of correlations for each correlation type.
Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial.

```{r ch040-biNegBinPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.align='center', fig.width = 8, fig.height=5,fig.cap = "`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
dat <- readRDS("./results/biNB_sims.rds")
dat %>%
    filter(cores == cores) %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    ## ggplot(aes(rho, rho_hat)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ + N) + theme_bw()
    facet_wrap(~ type + N) + theme_bw()
## ggsave('fig/plot-biNegBinPlot.pdf')
```
